\(Q\)-Boson model and relations with integrable hierarchies

### 2024-06-22 referee-answer.txt

Dear Editor,

Below are my responses to the referee's assessment:

Dear Referee,

I agree that my results are somewhat buried in the manuscript. Let me
start by saying that I was very careful not to claim credit for any
previous derivations. Some discussions in my current paper have been
addressed before, but not clearly enough. Therefore, this general
exposition also highlights my work and the field in
general. Additionally, there is pedagogical value in this.

Now, let me explain my main new results.

Section 3:

This section has more overlap with existing literature, particularly
with a paper by Wheeler that I cite in my work. In his paper, Wheeler
shows that the inner product of two Bethe states are KP hierarchy tau
functions. He finds the coefficients of the expansion and demonstrates
that these coefficients satisfy Plücker relations. My Section 3.1
improves on this result. I argue that besides the KP hierarchy, these
scalar products are also Toda hierarchy tau functions, as they can be
expanded as a Cauchy identity, with the coefficients being Schur
functions (see equation (28)). If we fix one set of coefficients in
this expression, Wheeler's results are recovered, as argued in
equation (40).

Section 3.2 is new and considers correlation functions. I wanted to
determine when the correlation functions satisfy integrable hierarchy
equations. Using expressions derived by Bogoliubov (see eq. (42.a)), I
rewrote them in the appropriate form to check if the correlation
functions are KP/Toda tau functions. For example, (42.b) cannot be
written as a Toda tau function, but with some adjustments, it can be
written as (51), which is a KP tau function. From the expansion (52),
also derived by Bogoliubov, we find (56) and (57), showing that these
expressions generally do not satisfy integrable hierarchy equations.

Section 3.3 presents my main results. Here, I argue that the general
form of tau functions associated with this phase model must be (67),
where G is given by (63). I also mention that from this general
expression, we can consider interesting cases, such as (68), which
fall into the class of hypergeometric tau functions. These problems
are currently under investigation in my follow-up paper.

Section 4:

This section attempts to replicate the analysis done in Section
3. There aren't many known results in the literature regarding the
relation between these objects and KP/Toda tau functions. I use known
formulas to see if they are tau functions. Wheeler has shown that
scalar products can be viewed as KP tau functions, as before.

I then write these formulas in terms of simpler orthogonal
polynomials. Firstly, I show that these scalar products cannot be
naively written as Toda tau functions (see (84)). However, using some
known expansions (see (85)), we can expand the inner products in
different forms. In one of them, we define a new set of coordinates
(equation (102)), making the scalar product of Q-Bosons Toda tau
functions with respect to these new coordinates. This is my main
result in this section. Additionally, I propose a conjecture worth
investigating in the future, as mentioned in Remark 2, regarding the
existence of 'functional' Plücker coordinates in the infinite
Grassmannian.

My Remark 3 also presents a new result. I emphasize that I do not wish
to take undue credit. Reference [NR19] uses an intricate construction
to show that the Big Schur functions are KP tau functions. Although
significant, my results trivialize their analysis. Since a simple
coordinate transformation makes the Big Schur functions ordinary Schur
functions (see equation (101)), this fact is trivial. I have included
this as a remark rather than stating it bluntly.

Having said that, let me return to my previous assessment of other
references. I am very interested in this field but have noticed that
papers on these topics are scattered. For example, in section 3.1.1,
the relation between the phase model and the ensemble of 2D Coulomb
particles seems obvious, but there are no references. I hope this
paper will draw attention to this particular problem. As I mentioned,
I am currently exploring this topic, but it is more challenging than I
previously thought.

Overall, I do not view this work as a full collection of new results
but as setting the stage for a research field that is somewhat
inactive. I use previous results to organize and draw attention to
open problems. In this process, I prove some new results, which,
though not major, serve as a starting point to investigate those open
problems.

I have added comments to my conclusions and improved my introduction.

I hope these clarifications are helpful.

Best regards,
Thiago

### 2024-06-23 referee-answer.txt

Section 3.3 presents my main results. Here, I argue that the general
form of tau functions associated with this phase model must be (67),
where G is given by (63). I also mention that from this general
expression, we can consider interesting cases, such as (68), which
fall into the class of hypergeometric tau functions. These problems
are currently under investigation in my follow-up paper.

Section 4:

This section attempts to replicate the analysis done in Section
3. There aren't many known results in the literature regarding the
relation between these objects and KP/Toda tau functions. I use known
formulas to see if they are tau functions. Wheeler has shown that
scalar products can be viewed as KP tau functions, as before.

Observe that the determinant formula 83 is new. From this expression
we can conclude that these objects cannot be written as 'simple' Toda
tau functions.  I then write these formulas in terms of simpler
orthogonal polynomials. Using some known expansions (see (85)), we can
expand the inner products in different forms. In one of them, we
define a new set of coordinates (equation (102)), making the scalar
product of Q-Bosons Toda tau functions with respect to these new
coordinates. This is my main result in this section. Additionally, I
propose a conjecture worth investigating in the future, as mentioned
in Remark 2, regarding the existence of 'functional' Plücker
coordinates in the infinite Grassmannian.

@@

My Remark 3 also presents a new result. I emphasize that I do not wish
to take undue credit. Reference [NR19] uses an intricate construction
to show that the Big Schur functions are KP tau functions. Although
significant, my results trivialize their analysis. Since a simple
coordinate transformation makes the Big Schur functions ordinary Schur
functions (see equation (101)), this fact is trivial. I have included
this as a remark rather than stating it bluntly.

Having said that, let me return to my previous assessment of other
references. I am very interested in this field but have noticed that
papers on these topics are scattered. For example, in section 3.1.1,
the relation between the phase model and the ensemble of 2D Coulomb
particles seems obvious, but there are no references. I hope this
paper will draw attention to this particular problem. As I mentioned,
I am currently exploring this topic, but it is more challenging than I
previously thought.

Overall, I do not view this work as a full collection of new results
but as setting the stage for a research field that is somewhat
inactive. I use previous results to organize and draw attention to
open problems. In this process, I prove some new results, which,
though not major, serve as a starting point to investigate those open
problems.

I have added these comments to my conclusions and I have also improved
my introduction.

### 2024-06-23 draft.tex
improve text

The exploration of connections between quantum and classical
exactly solvable models constitutes a significant endeavor aimed at
elucidating the overarching structure of integrable systems.  This
research avenue has yielded fruitful insights, as evidenced
by~\cite{Its:1992bj, Foda:2009zz, Alexandrov:2011aa, Araujo:2021ghu},
to name a few.  The present work is situated within this research field.

Here, we examine the emergence of classical integrable structures
within correlation functions of the q-boson system. This quantum
integrable system describes q-deformed bosons confined to
a one-dimensional chain~\cite{Bogoliubov:1992, Bogoliubov:1997soj,
  Bogoliubov2005}. Notably, this model bears close relation to the AL
(Ablowitz-Ladik) model, which is an integrable discretization of the
nonlinear Schrödinger equation.

In~\cite{Bogoliubov2005}, the author demonstrates that the \(q = 0\)
limit of the q-boson, known as the \emph{phase model}, is associated
with the enumeration of plane partitions. This connection allows for a
Schur polynomials expansion for the Bethe states. Subsequently, this
result was extended to \(q > 0\) in~\cite{Tsilevich:2006}, where the
Bethe states are expressed in terms of Hall-Littlewood polynomials.

@@

In Section 2, we provide a review of the phase model and
q-bosons. This section serves to establish notation and emphasize the
key aspects relevant to our analysis.  Section 3 delves into the
analysis of the phase model and its connections with the KP and Toda
integrable hierarchies. We explore the general expansion of a phase
model correlation function that yields a Toda system solution.  In
Section 4, we shift our focus to the q-bosons. We discuss a
determinantal formula for the scalar product of two off-shell Bethe
states and its expansion in terms of big, supersymmetric Schur
functions, as well as in terms of Kostka-Foulkes polynomials.
Finally, in Section 5, we conclude with an overview of open problems.

@@

%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*
\section{Conclusions and Perspectives}
%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*

@@

In Section 3, we delved into some aspects of the phase model. We
presented the scalar product of two off-shell Bethe states as an
elementary example of the Toda system tau function. Additionally, we
revealed an intriguing alternative portrayal of these scalar products
using a matrix integral framework, which corresponds to an ensemble of
Coulomb particles. This discovery opens up promising avenues for
further inquiry. Specifically, it would be intriguing to further
explore this subject and investigate whether the matrix model
description provides valuable insights into the phase model and its
Bethe roots. Such an investigation promises to illuminate the
underlying dynamics of the phase model and its relationship with
classical integrable systems.

We have elucidated how various correlation functions align with this
framework, revealing a remarkably rich structure. Notably, we explored
a mapping between the phase model data and the vertex operator
representation of free bosons. An intriguing avenue for future
research is to delve into the properties of the hypergeometric tau
functions uncovered in Section 3.3.1.  Exploring these functions is
expected to provide significant understanding, and comparing them with
existing literature will further enrich our comprehension of their
characteristics and implications within the context of integrable
systems.

In Section 4, we addressed the same problem within the context of the
q-boson model. Initially, we derived a determinant formula for the
q-boson scalar products, followed by a discussion on their expression
in terms of the phase model data. This result establishes a connection
between the q-boson and phase model quantities. Furthermore, we
explored different expansions for these scalar products. Notably, we
observed that they can be expanded in terms of Big Schur and
supersymmetric tau functions. Hence, we introduced a new set of
coordinates, demonstrating that scalar products can be precisely
expressed as Toda tau functions within this transformed coordinate
system.

>>>>

The study of connections between quantum and classical
exactly solvable models is an important research program aimed at
elucidating the underlying structure of integrable systems. This
research program has yielded fruitful insights, as evidenced
by~\cite{Its:1992bj, Foda:2009zz, Alexandrov:2011aa, Araujo:2021ghu},
to name a few. The present work is situated within this research field.

Here, we examine the emergence of classical integrable structures
within correlation functions of the q-boson system. This quantum
integrable system describes q-deformed bosons confined to a
one-dimensional chain~\cite{Bogoliubov:1992, Bogoliubov:1997soj,
  Bogoliubov2005}.  Interestingly, this model is closely related to
the AL (Ablowitz-Ladik) model, which is an integrable discretization
of the nonlinear Schrödinger equation.

In~\cite{Bogoliubov2005}, the author shows that the \(q = 0\)
limit of the q-boson, known as the \emph{phase model}, is associated
with the enumeration of plane partitions. This connection allows for a
Schur polynomials expansion for the Bethe states. Subsequently, this
result was extended to \(q > 0\) in~\cite{Tsilevich:2006}, where the
Bethe states are expressed in terms of Hall-Littlewood polynomials.

@@

In Section 2, we provide a review of the phase model and
q-bosons. This section serves to establish notation and emphasize the
key aspects relevant to our analysis.  Section 3 delves into the
analysis of the phase model and its connections with the KP and Toda
integrable hierarchies. We explore the general expansion of a phase
model correlation function that yields a Toda system solution.  In
Section 4, we shift our focus to the q-bosons. We discuss a
determinantal formula for the scalar product of two off-shell Bethe
states and its expansion in terms of big, supersymmetric Schur
functions, as well as in terms of Kostka-Foulkes polynomials.
Finally, in Section 5, we conclude with a discussion of our results
and an overview of open problems.

@@

%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*
\section{Discussion, Conclusions and Perspectives}
%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*%*

@@

In Section 3, we explored various aspects of the phase model,
presenting the scalar product of two off-shell Bethe states as an
elementary example of a Toda system tau function. This section
overlaps with existing literature, particularly a paper by Wheeler and
Foda~\cite{Wheeler:2010vmq, Foda:2008hn}. In their work, they
demonstrate that the inner product of two Bethe states are KP
hierarchy tau functions. Our current work builds on this by arguing
that these scalar products are also Toda hierarchy tau functions, as
they can be expanded as a Cauchy identity, with the coefficients being
Schur functions. When one set of coefficients is fixed in this
expression, Wheeler's results are recovered.

We also extend this analysis to other correlation functions,
determining when these functions satisfy integrable hierarchy
equations. Using expressions derived by Bogoliubov and
collaborators~\cite{Bogoliubov:1997soj}, we reformulated them to check
if the correlation functions are KP/Toda tau functions. Although these
objects cannot generally be written as Toda tau functions, with some
adjustments, they can be expressed as KP tau functions.  We have
elucidated how various correlation functions align with this
framework, revealing a remarkably rich structure.

The main results are discussed in Section 3.3, where we examine the
general form of tau functions associated with this phase model.
Notably, we explored a mapping between the phase model
data and the vertex operator representation of free bosons. An
intriguing avenue for future research is to delve into the properties
of the hypergeometric tau functions uncovered in Section 3.3.1.
Exploring these functions is expected to provide significant
understanding, and comparing them with existing literature will
further enrich our comprehension of their characteristics and
implications within the context of integrable systems. This problem is
currently under investigation.

Additionally, we revealed an intriguing alternative portrayal of these
scalar products using a matrix integral framework, which corresponds
to an ensemble of Coulomb particles. This discovery opens up promising
avenues for further inquiry. Specifically, it would be intriguing to
further explore this subject and investigate whether the matrix model
description provides valuable insights into the phase model and its
Bethe roots. Such an investigation promises to illuminate the
underlying dynamics of the phase model and its relationship with
classical integrable systems.

In Section 4, we addressed the same problem within the context of the
q-boson model, attempting to replicate the analysis done in Section
3. There are few known results in the literature regarding the
relationship between these objects and KP/Toda tau functions, so we
used established formulas to investigate if the scalar products are
also tau functions.

Initially, we derived a determinant formula for the q-boson scalar
products and discussed their expression in terms of the phase model
data. This result establishes a connection between the q-boson and
phase model quantities. Furthermore, we explored different expansions
for these scalar products. Notably, we observed that they can be
expanded in terms of Big Schur and supersymmetric tau
functions. Consequently, we introduced a new set of coordinates,
demonstrating that scalar products can be precisely expressed as Toda
tau functions within this transformed coordinate system.

We also wrote these formulas in terms of simpler orthogonal
polynomials. Firstly, we showed that these scalar products cannot be
naively written as Toda tau functions. However, using some known
expansions, we expanded the inner products in different forms. In one
of these expansions, we defined a new set of coordinates, making the
scalar product of q-bosons Toda tau functions with respect to these
new coordinates. This is the main result of this section.
